---
title: "PROJECT.data.exploration"
author: "Kendra Maas"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true # table of content true
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("ggplot2")
# install.packages("vegan")
# install.packages("ecodist")
# install.packages("RColorBrewer")
# install.packages("tidyr")
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("indicspecies")
# install.packages("devtools")
install.packages("vegan3d")

library(ggplot2)
library(vegan)
library(ecodist)
library(RColorBrewer)
library(tidyr)
library(tidyverse)
library(reshape2)
library(plyr)
library(indicspecies)
library(devtools)
library(vegan3d)
# devtools::install_github("AckerDWM/gg3D")

# library(gg3D)

ggplotRegression <- function (fit) {
    
    # require(ggplot2)
    
    ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
        geom_point() +
        stat_smooth(method = "lm", col = "red") +
        labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                           "Intercept =",signif(fit$coef[[1]],5 ),
                           " Slope =",signif(fit$coef[[2]], 5),
                           " P =",signif(summary(fit)$coef[2,4], 5)))
}

parseDistanceDF = function(phylip_file) {

   # Read the first line of the phylip file to find out how many sequences/samples it contains
    temp_connection = file(phylip_file, 'r')
    len = readLines(temp_connection, n=1)
    len = as.numeric(len)
    len = len +1
    close(temp_connection)
   
   
    phylip_data = read.table(phylip_file, fill=T, row.names=1, skip=1, col.names=1:len)
    colnames(phylip_data) <- row.names(phylip_data)
    return(phylip_data)
}




```

##Read in data

```{r read in data}

    otu <- read.table(file ="../PROJECT.trim.contigs.good.unique.good.filter.precluster.denovo.vsearch.pick.opti_mcc.0.03.subsample.shared", header=T, stringsAsFactors = FALSE, row.names=2)
otu <- select(otu, -label, -numOtus)


taxa <- read.table(textConnection(gsub("\\(.+?\\);", "\t", readLines("../PROJECT.trim.contigs.good.unique.good.filter.precluster.denovo.vsearch.pick.opti_mcc.0.03.cons.taxonomy"))), col.names=c("OTU", "Size", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus"), skip=1)
taxa <- taxa[taxa$OTU %in% names(otu),]

# get OTU abundance for this subsampling
sub.size <- data.frame(OTU = colnames(otu), size.sub = colSums(otu))

taxa <- full_join(taxa, sub.size, by = "OTU", copy=TRUE)

# selected taxa for stacked bars, you have to manupilate tax.summary to create this file
taxa.bars <- read.table(file="../PROJECT.taxabars.csv", header=T, sep=",")



maxab <- apply(otu, 2, max)
n1 <- names(which(maxab < 50))
otu.ab <- otu[,-which(names(otu) %in% n1)]
taxa.ab <- taxa[-which(taxa$OTU %in% n1),]



# Alpha diversity

alpha <- read.table(file="../PROJECT.trim.contigs.good.unique.good.filter.precluster.denovo.vsearch.pick.opti_mcc.groups.ave-std.summary", header=T, stringsAsFactors = FALSE)
alpha <- filter(alpha, label== "0.03" & method == "ave")

# Beta diversity
jc <- parseDistanceDF("../PROJECT.trim.contigs.good.unique.good.filter.precluster.denovo.vsearch.pick.opti_mcc.jest.0.03.lt.ave.dist")
bc <- parseDistanceDF("../PROJECT.trim.contigs.good.unique.good.filter.precluster.denovo.vsearch.pick.opti_mcc.braycurtis.0.03.lt.ave.dist")
tyc <- parseDistanceDF("../PROJECT.trim.contigs.good.unique.good.filter.precluster.denovo.vsearch.pick.opti_mcc.thetayc.0.03.lt.ave.dist")

# experimental data #
expdata <- read.table(file="../PROJECTenvData.csv", sep=",", header=T, stringsAsFactors = TRUE)


alpha.expdata <- left_join(alpha, expdata, by="group")


zymo <- alpha.expdata$treatment != "control"
alpha.expdata <- alpha.expdata[zymo,]
alpha.expdata <- droplevels.data.frame(alpha.expdata)
jc <- jc[zymo, zymo]
bc <- bc[zymo, zymo]
tyc <- tyc[zymo, zymo]
otu <- otu[zymo,]
taxa.bars <- taxa.bars[zymo,]

taxa.bars <- left_join(taxa.bars, expdata, by ="group")
taxa.bars$treatment <- factor(taxa.bars$treatment)


```

##Alpha Diversity
Boxplots of alpha diversity measures (diversity within a sample)

I'm adding the actual data points to the first one just to demonstrate the possiblility. Sometimes I like it in addition to the boxplot

```{r alpha diversity Treatment}
# alpha diversity (diversity within a sample)

xlab <- paste(levels(as.factor(alpha.expdata$TREATMENT)),"\n(N=", table(alpha.expdata$TREATMENT),")", sep="")

alpha.expdata %>%
ggplot( (aes(x=TREATMENT, y=sobs)))+
    geom_boxplot(varwidth = T, outlier.shape = NA)+
    geom_jitter(width=0.2)+
    scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial Richness by treatment")+
    labs(y="Species Observed", x="")
ggsave(file="richnessTREATMENT.jpg")

alpha.expdata %>%
ggplot( (aes(x=TREATMENT, y=sobs)))+
    geom_boxplot(varwidth = T, outlier.shape = NA)+
    geom_jitter(width=0.2)+
    facet_grid(.~TREATMENT2)+
    # scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial Richness by treatment")+
    labs(y="Species Observed", x="")
ggsave(file="richnesstreatment.jpg")


xlab <- paste(levels(as.factor(alpha.expdata$TREATMENT)),"\n(N=", table(alpha.expdata$TREATMENT),")", sep="")

alpha.expdata %>%
ggplot( (aes(x=TREATMENT, y=shannon)))+
    geom_boxplot(varwidth = T, outlier.shape = NA)+
    geom_jitter(width=0.2)+
    # facet_grid(.~TREATMENT2)+
    scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial diversity by treatment")+
    labs(y="Shannon", x="")
ggsave(file="shannonTREATMENT.jpg")


alpha.expdata %>%
ggplot( (aes(x=TREATMENT, y=shannon)))+
    geom_boxplot(varwidth = T, outlier.shape = NA)+
    geom_jitter(width=0.2)+
    facet_grid(.~TREATMENT2)+
    # scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial diversity by treatment")+
    labs(y="Shannon", x="")
ggsave(file="shannontreatment.jpg")


res.aov <- aov(shannon~TREATMENT, data=alpha.expdata)
summary(res.aov)
TukeyHSD(res.aov)



xlab <- paste(levels(as.factor(alpha.expdata$TREATMENT)),"\n(N=", table(alpha.expdata$TREATMENT),")", sep="")

alpha.expdata %>%
ggplot( (aes(x=TREATMENT, y=invsimpson)))+
    geom_boxplot(varwidth = T, outlier.shape = NA)+
    geom_jitter(width=0.2)+
    # facet_grid(.~TREATMENT2)+
    scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial diversity by treatment")+
    labs(y="Inverse Simpson", x="")
ggsave(file="simpsonTREATMENT.jpg")

alpha.expdata %>%
ggplot( (aes(x=TREATMENT, y=invsimpson)))+
    geom_boxplot(varwidth = T, outlier.shape = NA)+
    geom_jitter(width=0.2)+
    facet_grid(.~TREATMENT2)+
    scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial diversity by treatment")+
    labs(y="Inverse Simpson", x="")
ggsave(file="simpsontreatment.jpg")



res.aov <- aov(invsimpson~TREATMENT, data=alpha.expdata)
summary(res.aov)
TukeyHSD(res.aov)

```

##Beta Diversity 

Non-metric multidimensional scaling to visualize between sample diversity

###jc= Jaccard dissimilarity measure presence absence
```{r beta div nms jc}

jc.nms <- metaMDS(as.dist(jc), k=2, trymin=50, trymax=500, wascores=F, trace=FALSE)
jc.points <- data.frame(jc.nms$points)
jc.plot <- ggplot(jc.points, aes(x=MDS1, y=MDS2,label=rownames(jc)))

x <- max(jc.points$MDS1)/1.5
y <- min(jc.points$MDS2)


jc.plot +  geom_point(aes(shape = factor(alpha.expdata$TREATMENT2), fill=factor(alpha.expdata$TREATMENT)), size=5)+
    # geom_text()+
    scale_shape_manual(values = c(24, 21))+
    scale_fill_brewer(palette = "YlGnBu")+
    # scale_fill_manual(values=treat.col)+
    theme_bw()+
    annotate("text", x, y, label=paste("stress = ", round(jc.nms$stress, digits = 3)))+
    labs(x="", y="")+
    ggtitle("Jaccard")+
    guides(fill=guide_legend(override.aes = list(shape=21)))
ggsave(file="jc.treatment.nms.jpg")



```

###bc= Bray-Curtis dissimilarity measure that incoorporates abundance
```{r beta div nms bc}

## Whole dataset

bc.nms <- metaMDS(as.dist(bc), k=2, try=50, trymax=500, wascores=F, trace=FALSE)
bc.points <- data.frame(bc.nms$points)
bc.plot <- ggplot(bc.points, aes(x=MDS1, y=MDS2,label=rownames(bc)))


x <- max(bc.points$MDS1)/1.5
y <- min(bc.points$MDS2)


bc.plot +   geom_point(aes(shape = factor(alpha.expdata$TREATMENT2), fill=factor(alpha.expdata$TREATMENT)), size=5)+
    theme_void()+
    # axes_3D()+
    # stat_3D()+
    # geom_text()+
    scale_shape_manual(values = c(24, 21), name="TREATMENT2",labels = c("TREATMENT2", "Unchallenged"))+
    scale_fill_brewer(palette = "YlGnBu", name = "Linalool Levels")+
    # scale_fill_manual(values=treat.col)+
    stat_ellipse(aes(color=alpha.expdata$TREATMENT2))+
    scale_color_manual(values=c("black", "black"), , guide = 'none')+
    theme_bw()+
    annotate("text", x, y, label=paste("stress = ", round(bc.nms$stress, digits = 3)))+
    labs(x="", y="")+
    guides(fill=guide_legend(override.aes = list(shape=21)))+
     ggtitle("Bray Curtis")

ggsave(file="bc.treatment.nms.jpg")



```



###tyc= Theta YC dissimilarity measure that incoorporates abundance
```{r beta div nms tyc}

## Whole dataset

tyc.nms <- metaMDS(as.dist(tyc), k=2, trymin=50, trymax=500, wascores=F, trace=FALSE)
tyc.points <- data.frame(tyc.nms$points)
tyc.plot <- ggplot(tyc.points, aes(x=MDS1, y=MDS2,label=rownames(tyc)))

x <- max(tyc.points$MDS1)/1.5
y <- min(tyc.points$MDS2)


tyc.plot +    geom_point(aes(shape = factor(alpha.expdata$TREATMENT2), fill=factor(alpha.expdata$TREATMENT)), size=5)+
    # geom_text()+
    scale_shape_manual(values = c(24, 21), name="TREATMENT2",labels = c("TREATMENT2", "Unchallenged"))+
    scale_fill_brewer(palette = "YlGnBu", name = "Linalool Levels")+
    # scale_fill_manual(values=treat.col)+
    stat_ellipse(aes(color=alpha.expdata$TREATMENT2))+
    scale_color_manual(values=c("black", "black"), , guide = 'none')+
    theme_bw()+
    annotate("text", x, y, label=paste("stress = ", round(tyc.nms$stress, digits = 3)))+
    labs(x="", y="")+
    guides(fill=guide_legend(override.aes = list(shape=21)))+
     ggtitle("Theta YC")

ggsave(file="tyc.treatment.nms.jpg")



```

###Permanova
```{r beta div permanova}

permanova <- adonis2(as.dist(jc)~alpha.expdata$TREATMENT2 *alpha.expdata$TREATMENT, perm=999, rm.na=TRUE)
permanova


permanova <- adonis2(as.dist(bc)~alpha.expdata$TREATMENT2*alpha.expdata$TREATMENT, perm=999, rm.na=TRUE)
permanova


permanova <- adonis2(as.dist(tyc)~alpha.expdata$treatment, perm=999, rm.na=TRUE)
permanova


```


##Indicator species No control

Method for pulling out the species that coorespond with grouping variable (i.e. Treatment). See DeCaceres and Legendre
2009 for details.

this is failing, will troubleshoot in the future

```{r indicator species}
# 
# # otu[,-1] tells multipatt to ignore the first column which holds the name
# 
# 
# indic <- multipatt(otu[,-1], alpha.expdata$treatment,control = how(nperm=99))
# 
# summary(indic)
# 
# write.csv(file="indicator.species.csv",indic$sign%>%
#     rownames_to_column(var = "OTU")%>%
#     mutate(p.fdr = round(p.adjust(p.value, "fdr"),3))%>%
#     right_join(taxa, by = "OTU")%>%
#     # filter(p.fdr < 0.5) %>%
#     arrange(index))
# 
# 
# sig.otus <- indic$sign%>%
#     add_rownames(var="OTU")%>%
#     filter(p.adjust(p.value,"fdr") < 0.05)
# 
# #### Only very low abundance OTUs were pulled as indicators
# 
# # otu.indic <- otu[,which(names(otu) %in% sig.otus$OTU)]
# # taxa.indic <- taxa[which(taxa$OTU %in% sig.otus$OTU),]
```


###Taxa Barplot
```{r Barplot, fig.height=10, fig.width=8}

 taxa.col <- c(
     "Firmicutes_other"="#ff7f00", #orange brown
     "Lactobacillus"="#fdbf6f", #orange
     "Bifidobacteriaceae"="#79C360", #green
     "Pseudomonadaceae"="#A6CEE3", #lightest blue
     "Lactobacillus"="#7DB4D5", #second lightest blue
     "Enterobacteriaceae_other"="#08519c", # darkest blue
     "Salmonella"="#5C9FC9", #middle blue
     "Clostridia"="#3A89BD", #second darkest blue
     "Proteobacteria_other" = "#043363", # really dark blue
     "Actinobacteriota_other"="#4a773a", #dark green
     "Escherichia-Shigella"="#fdae6b", #light orange
     "Bacteria_other"="#d3d3d3", #grey
     "Bacteroidetes"="#6a3d9a") #black


 tax <- melt(taxa.bars, id.vars = c("group", "treatment", "TREATMENT2",  "TREATMENT3", "TREATMENT4", "day" ))
 
 
 
 ##check that melt occured the way you expected and that factors are factors
 str(tax)

 
 
 tax2 <- aggregate(value~variable+treatment+TREATMENT2, data=tax, FUN=sum)

 

ggplot(tax2,  aes( y=value, x=factor(treatment), color=NULL, fill=factor(variable), order=-as.numeric(variable)))+ #order=- makes the stack the same order as the legend
   geom_bar(position="fill", stat="identity")+
     xlab("Treatment")+
     ylab("Percent total community")+
     facet_grid(TREATMENT2~.)+
     scale_fill_manual(values=taxa.col, guide=guide_legend(title="Taxonomic level"))+
     theme_bw()
 ggsave(file="barplot.faceted.jpg", height=5, width=5, dpi=600)

 
# separate out unchallenged
 tax2 %>%
     filter(TREATMENT2 == "Unchallenged")%>%
ggplot(  aes( y=value, x=factor(treatment), color=NULL, fill=factor(variable), order=-as.numeric(variable)))+ #order=- makes the stack the same order as the legend
   geom_bar(position="fill", stat="identity")+
     xlab("Treatment")+
     ylab("Percent total community")+
     scale_fill_manual(values=taxa.col, guide=guide_legend(title="Taxonomic level"))+
     theme_bw()
 ggsave(file="barplot.unchallenged.jpg", height=5, width=5, dpi=600)

# separate out challenged
 tax2 %>%
     filter(TREATMENT2 == "SE")%>%
ggplot(  aes( y=value, x=factor(treatment), color=NULL, fill=factor(variable), order=-as.numeric(variable)))+ #order=- makes the stack the same order as the legend
   geom_bar(position="fill", stat="identity")+
     xlab("Treatment")+
     ylab("Percent total community")+
     scale_fill_manual(values=taxa.col, guide=guide_legend(title="Taxonomic level"))+
     theme_bw()
 ggsave(file="barplot.challenged.jpg", height=5, width=5, dpi=600)

 
 #add in day
 tax2 <- aggregate(value~variable+treatment+TREATMENT2+day, data=tax, FUN=sum)

 

ggplot(tax2,  aes( y=value, x=factor(TREATMENT2), color=NULL, fill=factor(variable), order=-as.numeric(variable)))+ #order=- makes the stack the same order as the legend
   geom_bar(position="fill", stat="identity")+
     xlab("Treatment")+
     ylab("Percent total community")+
     facet_grid(treatment~day)+
     scale_fill_manual(values=taxa.col, guide=guide_legend(title="Taxonomic level"))+
     theme_bw()
 ggsave(file="barplot.faceted.day.treatment.jpg", height=5, width=5, dpi=600)

 
# separate out unchallenged
 tax2 %>%
     filter(TREATMENT2 == "Unchallenged")%>%
ggplot(  aes( y=value, x=factor(day), color=NULL, fill=factor(variable), order=-as.numeric(variable)))+ #order=- makes the stack the same order as the legend
   geom_bar(position="fill", stat="identity")+
     xlab("Treatment")+
          facet_grid(.~treatment)+
     ylab("Percent total community")+
     scale_fill_manual(values=taxa.col, guide=guide_legend(title="Taxonomic level"))+
     theme_bw()
 ggsave(file="barplot.unchallenged.day.jpg", height=5, width=5, dpi=600)
 
 # separate out challenged
 tax2 %>%
     filter(TREATMENT2 == "SE")%>%
ggplot(  aes( y=value, x=factor(day), color=NULL, fill=factor(variable), order=-as.numeric(variable)))+ #order=- makes the stack the same order as the legend
   geom_bar(position="fill", stat="identity")+
     xlab("Treatment")+
          facet_grid(.~treatment)+
     ylab("Percent total community")+
     scale_fill_manual(values=taxa.col, guide=guide_legend(title="Taxonomic level"))+
     theme_bw()
 ggsave(file="barplot.challenged.day.jpg", height=5, width=5, dpi=600)

```



## radEmu for pulling OTUs that are significantly different between treatments 

```{r radEmu}
# install.packages("devtools")
# devtools::install_github("statdivlab/radEmu")
# library(radEmu)

maxab <- apply(otu, 2, max)
n1 <- names(which(maxab < 50))
otu.ab <- otu[,-which(names(otu) %in% n1)]

row.names(alpha.expdata) <- alpha.expdata$group



all_fit <- emuFit(formula = ~ Treatment,
                  data = alpha.expdata,
                  Y = otu.ab)

all_fit$coef %>%
    filter(pval <0.1)

# compare high vs everything else
alpha.expdata$treat2 <- as.character(alpha.expdata$Treatment)
#edited to change Control, Medium, Low to "else"
fix(alpha.expdata)



all_fit <- emuFit(formula = ~ treat2,
                  data = alpha.expdata,
                  Y = otu.ab.denit)

write.csv(file = "radEmuAlpha0.05.csv", all_fit$coef %>%
    filter(pval <0.05) %>%
    left_join(taxa, by = c("category"= "OTU") ))



```                  

