---
title: "genericFirstLook"
author: "Kendra Maas"
date: "Sept , 2016"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("ggplot2")
# install.packages("vegan")
# install.packages("ecodist")
# install.packages("RColorBrewer")
# install.packages("tidyr")
# install.packages("dplyr")
# install.packages("knitr")
# install.packages("indicspecies")
library(ggplot2)
library(vegan)
library(ecodist)
library(RColorBrewer)
library(tidyr)
library(dplyr)
library(indicspecies)

ggplotRegression <- function (fit) {
    
    require(ggplot2)
    
    ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
        geom_point() +
        stat_smooth(method = "lm", col = "red") +
        labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                           "Intercept =",signif(fit$coef[[1]],5 ),
                           " Slope =",signif(fit$coef[[2]], 5),
                           " P =",signif(summary(fit)$coef[2,4], 5)))
}

parseDistanceDF = function(phylip_file) {

   # Read the first line of the phylip file to find out how many sequences/samples it contains
    temp_connection = file(phylip_file, 'r')
    len = readLines(temp_connection, n=1)
    len = as.numeric(len)
    len = len +1
    close(temp_connection)
   
   
    phylip_data = read.table(phylip_file, fill=T, row.names=1, skip=1, col.names=1:len)
    colnames(phylip_data) <- row.names(phylip_data)
    return(phylip_data)
}




    ```

##R Markdown

```{r read in data}

otu <- read.table(file = "../YOURPROJECT.trim.contigs.good.unique.good.filter.precluster.pick.pick.an.unique_list.0.03.subsample.shared", header=T, stringsAsFactors = FALSE)
otu <- select(otu, -label, -numOtus)

# only file that need manupalition before reading in, this is based on cons.taxonomy
taxa <- read.csv(file="YOURPROJECT.0.03.taxa.csv", header=T, sep=",", stringsAsFactors = FALSE)
taxa <- taxa[taxa$OTU %in% names(otu),]

maxab <- apply(otu, 2, max)
n1 <- names(which(maxab < 500))
otu.ab <- otu[,-which(names(otu) %in% n1)]
taxa.ab <- taxa[-which(taxa$OTU %in% n1),]

# Alpha diversity

alpha <- read.table(file="../YOURPROJECT.trim.contigs.good.unique.good.filter.precluster.pick.pick.an.unique_list.groups.ave-std.summary", header=T, stringsAsFactors = FALSE)
alpha <- filter(alpha, label== "0.03" & method == "ave")

# Beta diversity
jc <- parseDistanceDF("../YOURPROJECT.trim.contigs.good.unique.good.filter.precluster.pick.pick.an.unique_list.jest.0.03.lt.ave.dist")
bc <- parseDistanceDF("../YOURPROJECT.trim.contigs.good.unique.good.filter.precluster.pick.pick.an.unique_list.braycurtis.0.03.lt.ave.dist")
tyc <- parseDistanceDF("YOURPROJECT.trim.contigs.good.unique.good.filter.precluster.pick.pick.an.unique_list.thetayc.0.03.lt.ave.dist")

# experimental data
expdata <- read.csv(file="experiment.csv", header=T, stringsAsFactors = FALSE)
expdata <- arrange(expdata, sample)

#I create a csv with client sample and group that has sequence name (Client name + MARS barcode)

group.dat <- read.csv(file="group.csv", header=T, stringsAsFactors = FALSE)

expdata <- left_join(expdata, group.dat, on="sample")

alpha.expdata <- left_join(alpha, expdata, on="group")




```

Boxplots of alpha diversity measures (diversity within a sample)

I'm adding the actual data points to the first one just to demonstrate the possiblility. Sometimes I like it in addition to the boxplot

```{r alpha diversity treatment}
# alpha diversity (diversity within a sample)
xlab <- paste(levels(as.factor(alpha.expdata$treatment)),"\n(N=", table(alpha.expdata$treatment),")", sep="")
simp <-  ggplot(data=alpha.expdata, (aes(x=treatment, y=invsimpson)))+
    geom_boxplot(varwidth = T)+
    geom_jitter(width=0.2)+
    scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial Diversity by Treatment")

simp
```

Linear Regression of the factors measured against Bacterial Diversity
```{r alpha diversity experimental measurements}




ggplotRegression(lm(invsimpson~YOURFACTOR, data = alpha.expdata))
ggplotRegression(lm(invsimpson~YOURFACTOR, data = alpha.expdata))

```

Linear Regression for each treatment of the factors measured against Bacterial Diversity
```{r alpha diversity experimental measurements by treatment}
ggplot(data=alpha.expdata, (aes(x=YOURFACTOR, y=invsimpson,color=treatment)))+
    geom_point(aes(size=3))+
    scale_color_brewer(palette="Spectral")+
    stat_smooth(method="lm")+
    theme_bw()+
    ggtitle("Bacterial Diversity by YOURFACTOR")



```

=======
jc=Jaccard dissimilarity measure, presence absence. This one overweights rare species by counting a rare species and an abundant on as simply present
```{r beta div nms j}

jc.nms <- metaMDS(as.dist(jc), k=2, trymin=50, trymax=500, wascores=F)
jc.points <- data.frame(jc.nms$points)
jc.plot <- ggplot(jc.points, aes(x=MDS1, y=MDS2))


jc.plot + geom_point(aes(color=factor(alpha.expdata$YOURFACTOR1),  size=alpha.expdata$YOURFACTOR2))+
    theme_bw()+
    scale_color_brewer(palette="Spectral")
```

bc= Bray-Curtis dissimilarity measure that incoorporates abundance
```{r beta div nms bc}

bc.nms <- metaMDS(as.dist(bc), k=2, trymin=50, trymax=500, wascores=F)
bc.points <- data.frame(bc.nms$points)
bc.plot <- ggplot(bc.points, aes(x=MDS1, y=MDS2))


bc.plot + geom_point(aes(color=factor(alpha.expdata$YOURFACTOR1),  size=alpha.expdata$YOURFACTOR2))+
    theme_bw()+
    scale_color_brewer(palette="Spectral")
```


tyc= Theta YC distance measure that strongly penalizes species that are abundant in one sample and missing in another. Strongly downweights rare species
```{r beta div nms tyc}

tyc.nms <- metaMDS(as.dist(tyc), k=2, trymin=50, trymax=250, wascores=F)
tyc.points <- data.frame(tyc.nms$points)
tyc.plot <- ggplot(tyc.points, aes(x=MDS1, y=MDS2))


tyc.plot + geom_point(aes(color=factor(alpha.expdata$YOURFACTOR1),  size=alpha.expdata$YOURFACTOR2))+
    theme_bw()+
    scale_color_brewer(palette="Spectral")


```


Hypothesis testing

Permanova results that I look at significance (Pr(>F)) and the R2 which can be interpreted as roughly the % varibility in your whole community that can be explained by the model factor 
```{r hypothesis testing}

permanova <- adonis(as.dist(jc)~alpha.expdata$YOURFACTOR, perm=999, rm.na=TRUE)
permanova

permanova <- adonis(as.dist(bc)~alpha.expdata$YOURFACTOR, perm=999, rm.na=TRUE)
permanova

permanova <- adonis(as.dist(tyc)~alpha.expdata$YOURFACTOR, perm=999, rm.na=TRUE)
permanova

permanova <- adonis(as.dist(jc)~alpha.expdata$YOURFACTOR1+ alpha.expdata$YOURFACTOR2, perm=999, rm.na=TRUE)
permanova

permanova <- adonis(as.dist(bc)~alpha.expdata$YOURFACTOR1+ alpha.expdata$YOURFACTOR2, perm=999, rm.na=TRUE)
permanova

permanova <- adonis(as.dist(tyc)~alpha.expdata$YOURFACTOR1+ alpha.expdata$YOURFACTOR2, perm=999, rm.na=TRUE)
permanova
```

```{r indicator species}

# otu[,-1] tells multipatt to ignore the first column which holds the name
indic <- multipatt(otu[,-1], alpha.expdata$treatment,control = how(nperm=999), duleg=TRUE)
# summary(indic)

write.csv(file="indicator.species.csv",indic$sign%>%
    add_rownames(var = "OTU")%>%
    right_join(taxa, by = "OTU")%>%
    filter(p.value < 0.05) %>%
    arrange(index))



```

\pagebreak
This NMS (Theta YC) has the strongest associated OTUs mapped on top as vectors, it's very messy but gives you and idea of how the samples relate to OTUs.

```{r nms with vectors, echo=FALSE}

# nms <- metaMDS(otu, distance ="bray", k=2, trymax=1000, wascores=TRUE)

##try making full plot with just vegan and ggplot from http://stackoverflow.com/questions/14711470/plotting-envfit-vectors-vegan-package-in-ggplot2

scrs <- as.data.frame(scores(tyc.nms, display ='sites'))
scrs <- cbind(scrs,
              YOURFACTOR1=alpha.expdata$YOURFACTOR1,
              Simpson=alpha.expdata$invsimpson)
spec <- envfit(tyc.nms, otu, perm=99)

spp.scrs <- as.data.frame((scores(spec, display="vectors")))
spp.scrs <- cbind(spp.scrs, Species =rownames(spp.scrs))
spp.scrs <- cbind(spp.scrs, pVal=spec$vectors$pvals)
spp.scrs <- cbind(spp.scrs, Phyla=taxa$YOURTAXA)

###color vectors by phyla

ggplot(scrs)+
    coord_fixed()+
    geom_segment(data=spp.scrs[spp.scrs$pVal<0.011,],
                 aes(x=0, xend=NMDS1, y=0, yend=NMDS2, color = Phyla),
                 arrow=arrow(length=unit (0.25, "cm")), size=1.5)+
    # scale_color_manual(values=spp.scrs$phy.col[spp.scrs$pVal<0.025])+
    geom_point(mapping=aes(x=NMDS1, y=NMDS2, shape = as.factor(treatment)), size=2)+
    # geom_text(data=scrs, aes(x=NMDS1-.02, y=NMDS2-.02, label=row.names(scrs)), size=4)+
    # geom_text(data =spp.scrs[spp.scrs$pVal<0.005,], aes(x=NMDS1, y=NMDS2, label=Phyla), size=4)+
    theme_bw()
```


Same as above but only fitting abundant OTUs (only keeping OTUs that are at least 5% of at least one sample)

```{r nms with vectors only abundant, echo=FALSE}

# nms <- metaMDS(otu, distance ="bray", k=2, trymax=1000, wascores=TRUE)

##try making full plot with just vegan and ggplot from http://stackoverflow.com/questions/14711470/plotting-envfit-vectors-vegan-package-in-ggplot2


spec <- envfit(tyc.nms, otu.ab, perm=99)

spp.scrs <- as.data.frame((scores(spec, display="vectors")))
spp.scrs <- cbind(spp.scrs, Species =rownames(spp.scrs))
spp.scrs <- cbind(spp.scrs, pVal=spec$vectors$pvals)
spp.scrs <- cbind(spp.scrs, Phyla=taxa.ab$family)

###color vectors by phyla

ggplot(scrs)+
    coord_fixed()+
    geom_segment(data=spp.scrs[spp.scrs$pVal<0.011,],
                 aes(x=0, xend=NMDS1, y=0, yend=NMDS2, color = Phyla),
                 arrow=arrow(length=unit (0.25, "cm")), size=1.5)+
    # scale_color_manual(values=spp.scrs$phy.col[spp.scrs$pVal<0.025])+
    geom_point(mapping=aes(x=NMDS1, y=NMDS2, shape = as.factor(treatment)), size=2)+
    # geom_text(data=scrs, aes(x=NMDS1-.02, y=NMDS2-.02, label=row.names(scrs)), size=4)+
    # geom_text(data =spp.scrs[spp.scrs$pVal<0.005,], aes(x=NMDS1, y=NMDS2, label=Phyla), size=4)+
    theme_bw()
```